Ride at Your Own Risk: Insights into Toronto’s Bicycle Theft Hotspots

### 2024-11-12 paper/paper.Rmd
paper revised

# Mapping the Danger Zones: High-Risk Locations Across Toronto
Use a map visualization to highlight theft hotspots across Toronto, pointing out neighborhoods or transit hubs with the highest incidence rates.

- map

@@

Map
```{r, include = TRUE, warning = FALSE, message = FALSE}
# Load cleaned coordinates data
coordinates_data <- read.csv(here("data/cleaned_coordinates_data.csv"))

# Plot the map with bike theft locations data
ggplot() +
  geom_sf(data = streets, color = "gray", size = 0.3, alpha = 0.6) + # OSM street map layer
  geom_point(
    data = coordinates_data, aes(x = longitude, y = latitude),
    size = 0.5, color = "red"
  ) + # Bike theft points
  labs(
    title = "Map of Bicycle Theft Locations in Toronto",
    x = "Longitude", y = "Latitude"
  ) +
  coord_sf(
    xlim = c(-79.6303856025883, -79.1180336515019),
    ylim = c(43.5822069220627, 43.8554571861712), expand = FALSE
  ) +
  theme_minimal() +
  theme(
    panel.background = element_blank(), # Remove panel background
    panel.grid.major = element_blank(), # Remove major grid lines
    panel.grid.minor = element_blank(), # Remove minor grid lines
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

>>>>

# Mapping the Danger Zones: High-Risk Locations Across Toronto

# Mapping the Danger Zones: High-Risk Locations Across Toronto

Now imagine you’re navigating the streets of Toronto, trying to understand where bicycle thefts are happening. The first chart is like an overhead view from a hot air balloon, showing clusters of incidents with bright orange circles scattered across the landscape. Each circle represents a collection of theft reports—smaller bubbles indicate fewer incidents, while larger ones stand like towering beacons, signaling areas of greater vulnerability. Downtown Toronto, unsurprisingly, is covered by a massive bubble, indicating that this densely populated urban center has been most affected by bike thefts. As we drift north and away from the downtown core, the bubbles shrink, suggesting fewer incidents—likely due to the suburban nature of these areas, where bikes are stored more securely. And here's the interactive part—you can click on these bubbles to zoom in and explore specific areas, uncovering the exact locations and getting a more detailed sense of bike stolen cases. It's like having a personal guide leading you through the hot spots and quieter neighborhoods.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
#### Map with Markers ####

# Load necessary libraries
library(dplyr)         # Provides functions for data manipulation.
library(leaflet)       # Enables interactive maps.
library(purrr)         # Contains functional programming tools, used here to walk through data.
library(readr)         # Provides functions to read rectangular data like CSV files.
library(here)          # Helps manage file paths in a more robust way.

# Load your dataset (adjust the file path accordingly)
map_data <- read_csv(here("data", "map_by_neighbourhoods_data.csv"), show_col_types = FALSE)
# Loads the CSV file from the specified path using the 'here()' function for consistent path management.
# 'show_col_types = FALSE' suppresses messages about the data types of each column.

# Modify the data frame, categorizing case_num column
map_data <- map_data %>% 
  mutate(
    mag.level = cut(case_num, c(100, 500, 1000, 1500),
                    labels = c('> 100 & <=500', '>500 & <=1000', '>1000 & <=1500'))
    # Creates a 'mag.level' column that categorizes 'case_num' into three levels based on specified ranges.
  ) %>%
  split(.$mag.level)  # Splits the dataset into a list of data frames by 'mag.level'.

# Initialize the leaflet map
l <- leaflet() %>%
  addProviderTiles(providers$Esri.OceanBasemap)
# Initializes a Leaflet map object and adds a base tile layer from the Esri Ocean Basemap provider.

# Add the markers to the map for each case_num level
names(map_data) %>%
  walk(function(df) {
    l <<- l %>% 
      addMarkers(
        data = map_data[[df]], lng = ~longitude, lat = ~latitude,
        label = ~paste0("Hood: ", hood_name, " | Stolen Place: ", stolen_place, " | Cost: ", cost, " | Status: ", status),
        popup = ~paste0(
          "Hood: ", hood_name, "<br>",
          "Stolen Place: ", stolen_place, "<br>",
          "Cost: ", cost, "<br>",
          "Status: ", status, "<br>",
          "Case in area: ", case_num
        ),
        group = df,
        clusterOptions = markerClusterOptions()
      )
  })
# Iterates over the names of the 'map_data' list (which are the different 'mag.level' categories).
# For each category, it adds markers to the map:
# - `data = map_data[[df]]`: Uses the split data for the current case_num level.
# - `lng = ~longitude, lat = ~latitude`: Sets longitude and latitude for marker placement.
# - `label`: Displays the 'case_num' value when markers are hovered over.
# - `popup`: Displays additional information (hood_num, stolen_place, cost, status, and case_num) when a marker is clicked.
# - `group = df`: Groups the markers by 'mag.level'.
# - `clusterOptions = markerClusterOptions()`: Enables clustering of markers for better visualization.

# Add layer controls and a minimap
l <- l %>%
  addLayersControl(
    overlayGroups = names(map_data),
    options = layersControlOptions(collapsed = FALSE)
  ) %>%
  addMiniMap(tiles = providers$Esri.OceanBasemap, width = 120, height = 80)
# Adds layer controls to allow switching between different 'mag.level' categories on the map.
# Adds a minimap showing a small overview of the area using the Esri Ocean Basemap.

# Display the Map
l
```

In the second chart, we dive deeper into the city with a ground-level view—a sea of red dots each marks an individual theft incident. The dots are packed tightly in certain neighborhoods, sprawling across downtown Toronto like a dense red blanket. It paints a vivid picture of the density of bike thefts—evidence that the bustling central part of the city faces a higher risk. Street after street is littered with these red markers, almost as if each streetlight bore witness to a theft. There’s a sharp concentration near transit hubs and popular neighborhoods, suggesting a correlation between foot traffic, bike use, and crime opportunity. The City has been making efforts to combat this issue; for example, Cycle Toronto launched the #EndBikeTheft campaign, advocating for secure parking and better registration systems, with city discussions exploring collaboration with platforms like 529 Garage to help reduce and recover stolen bikes (@citeEndBikeTheft). It’s almost as though you can hear the honks and feel the rush of people in these crowded streets.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Load necessary libraries
library(ggplot2)
library(sf)
library(osmdata)
library(here)

# Load cleaned coordinates data
coordinates_data <- read.csv(here("data/cleaned_coordinates_data.csv"))

# Plot the map with bike theft locations data
bike_theft_plot <- ggplot() +
  geom_sf(data = streets, color = "gray", size = 0.3, alpha = 0.6) + # OSM street map layer
  geom_point(
    data = coordinates_data, aes(x = longitude, y = latitude),
    size = 0.5, color = "red"
  ) + # Bike theft points
  labs(
    title = "Map of Stolen Bike Cases in Toronto",
    x = "Longitude", y = "Latitude"
  ) +
  coord_sf(
    xlim = c(-79.6303856025883, -79.1180336515019),
    ylim = c(43.5822069220627, 43.8554571861712), expand = FALSE
  ) +
  theme_minimal() +
  theme(
    panel.background = element_blank(), # Remove panel background
    panel.grid.major = element_blank(), # Remove major grid lines
    panel.grid.minor = element_blank(), # Remove minor grid lines
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

bike_theft_plot
```

Finally, the third chart takes a step back to help us understand the data by neighborhood, but this time with broader strokes. It presents different regions in Toronto as bubbles of varying sizes and colors, indicating the number of thefts in each. The larger and darker the bubble, the more incidents that have occurred there. One particular neighborhood stands out—a massive blue circle in the downtown area—it visually reinforces what the other charts have hinted at: that specific neighborhoods, especially those near the core, are the epicenters of bicycle theft. Meanwhile, the areas with smaller circles seem almost untouched in comparison, a stark contrast that reveals a city of disparities—some areas buzz with activity and theft risk, while others enjoy relative calm.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
#### Map with Areas ####

# Load necessary libraries
library(dplyr)
library(readr)
library(ggplot2)
library(plotly)
library(sf)
library(crosstalk)
library(here)
library(osmdata)

# Load your dataset (adjust the file path accordingly)
map_data <- read_csv(here("data", "map_by_area_data.csv"), show_col_types = FALSE)

# Fetch map data from OpenStreetMap (OSM)
toronto_map <- opq(bbox = toronto_bbox) %>%
  add_osm_feature(key = "boundary", value = "administrative") %>%
  osmdata_sf()

# Generate a similar plot using ggplot2 and plotly
shared_data <- crosstalk::SharedData$new(map_data)

# Create ggplot object with the dataset and Toronto map as a background
ggplot_map <- ggplot() +
  geom_sf(data = streets, color = "lightgray", size = 0.3, alpha = 0.6) +  # Add OSM street map layer
  geom_point(data = shared_data, aes(x = longitude, y = latitude, color = case_num, size = case_num, 
                                     text = paste("Total Cases: ", case_num, " | Area Name: ", area_name))) +
  coord_sf(
    xlim = c(-79.6303856025883, -79.1180336515019),
    ylim = c(43.5822069220627, 43.8554571861712), expand = FALSE
  ) +
  theme_minimal() +
  labs(title = "Toronto Stolen Bike Case Numbers by Area", x = "Longitude", y = "Latitude", color = "Number of Cases", size = "Total Cases") +
  theme(
    panel.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  scale_color_continuous(labels = scales::comma) +
  scale_size_continuous(range = c(3, 12)) # Adjust range values to control the size of points

# Convert ggplot object to interactive plotly object
plotly_map <- ggplotly(ggplot_map, tooltip = "text") %>%
  highlight(
    on = "plotly_hover",
    off = "plotly_doubleclick",
    selected = attrs_selected(line = list(color = "black"))
  )

plotly_map
```

@@

# Citation
